\documentclass[article,twocolumn]{memoir}
\usepackage[utf8]{inputenc}
\usepackage{psfrag}
\usepackage{hyperref}
\usepackage{color}
\usepackage[]{graphicx}

\graphicspath{{../figures/other/}{../figures/generated/}}


\title{Theatre in Mozambique}
\author{Rita Reis and Luis Pedro Coelho}

\bibliographystyle{alpha}
%\bibliographystyle{plain}
\pagestyle{Ruled}

\aliaspagestyle{chapter}{Ruled}
\makeatletter
\if@twoside
	\makeoddhead{Ruled}{}{}{\scshape\rightmark}
	\makeevenhead{Ruled}{\scshape\leftmark}{}{}
	\makeevenfoot{Ruled}{\pagenumberfont\thepage}{}{}
	\makeoddfoot{Ruled}{}{}{\pagenumberfont\thepage}
	% Put section number on top
	\def\sectionmark#1{\markright{#1 (\thesection)}}
	\def\chaptermark#1{\markboth{#1}{#1}}
	\renewcommand*{\bibmark}{\markboth{\bibname}{\bibname}} % I don't like empty headings!
\else

	% Put section number on top
	\def\sectionmark#1{\markright{#1 (\thesection)}}
	\def\chaptermark#1{\markright{#1 (\thechapter)}}

	\makeoddhead{Ruled}{}{}{\scshape\rightmark}
	\makeevenhead{Ruled}{}{}{\scshape\rightmark}
	\makeevenfoot{Ruled}{}{}{\pagenumberfont\thepage}
	\makeoddfoot{Ruled}{}{}{\pagenumberfont\thepage}
	\renewcommand*{\bibmark}{\markright{\bibname}} % I don't like empty headings!


\fi
\makeatother

%Change fonts: Page number sans-serif (much cleaner than the roman version)
\def\pagenumberfont{\sffamily}
% Change fonts: Section headers Sans-Serif:
\setsecheadstyle{\Large\sffamily\raggedright}
\setsubsecheadstyle{\large\sffamily\raggedright}
\setsubsubsecheadstyle{\normalsize\sffamily\raggedright}

% Title
\pretitle{\LARGE\sffamily}
\posttitle{\par\vspace{4ex}}
\preauthor{\large\sffamily\hspace{1cm}}
\postauthor{\par\vspace{3ex}}
\predate{\small\sffamily\hspace{1cm}Last Updated on: }
\postdate{\par\vspace{2cm}}
\copypagestyle{title}{plain}
\makeoddfoot{title}{}{}{}
\makeevenfoot{title}{}{}{}

\renewcommand{\abstractnamefont}{\sffamily}
\renewcommand{\abstracttextfont}{}
\renewcommand{\absnamepos}{flushleft}


\makechapterstyle{mestrado}{% Originally ``AlexanderGrebenkov'', adapted
\renewcommand{\chapterheadstart}{\goodbreak\vspace{\beforechapskip}}
\renewcommand{\chapnamefont}{\normalfont\Large\scshape}
\renewcommand{\chapnumfont}{\normalfont\Large\scshape}
\renewcommand{\chaptitlefont}{\normalfont\Large\scshape}
\renewcommand{\printchaptername}{}
\renewcommand{\chapternamenum}{}
\renewcommand{\printchapternum}{\normalfont\Large\scshape\S\thechapter}
\renewcommand{\afterchapternum}{\hspace{1em}}
\renewcommand{\afterchaptertitle}{\par\nobreak\vspace{-.9em}\moveright 6pt\vbox to 1pt{\hrule width .32\textwidth}\nobreak\vskip\afterchapskip\nobreak}
}
\chapterstyle{mestrado}

\begin{document}

\maketitle

\chapter{Introduction}
In July 2010, we worked with Kufunana, a group in Mozambique, which uses
theatre, particularly theatre of the oppressed, to reach out on HIV/AIDS
issues. We directed a show with the actors of the group, a show we developed
with them, loosely based on Arthur Schnitzler's play ``La Ronde.'' We also took
the opportunity to talk to the members of Kufunana and interview them about
their work more generally.

In Mozambique, particularly in the region we were in, the central region, HIV
is a major issue, with infection rates of circa 20\%.

In Mozambican society, theatre is regularly used by many groups to reach out
with their social messages. Kufunana is one of the best known groups in this
area, but it is representative of a wider phenomenon.

We will present and discuss both the work we did with them and what we learned
from them through both informal discussion and formal interviews.

\chapter{Kufunana's Work}

Through our informal discussions with the members of Kufunana as well as
structured interviews\footnote{We are making some of the videos available, with
English subtitles, online at: http://beiraproject.org.} we were able to gather
information on their actions in the social sphere.

Kufunana is a cultural organization that focuses of HIV/AIDS. Their main
artistic focus is theatrical, but they also have musicians, plastic artists,
and artisans. All of their members have been tested for HIV and some are
positive.

They often use theatre of the oppressed techniques in places around the city to
raise awareness on voluntary HIV testing (or sometimes other issues). They will
perform in markets, squares, discos, but also for private companies who hire
them to come and perform for their workers. This is actually one of their main
sources of income (which supports other activities). Companies have now
recognized that keeping their employees in good health is in their best
interests.

Their work will start with a short prepared play, but will intermingle with
discussion of the issues. The workers will often have the opportunity to get
tested immediately or receive counseling (the results are confidential, of
course).

Additionally, they often travel to the country-side to do their work. Their
typical approach is to combine theatrical performances, theatre of the
oppressed, pure entertainment, and discussions. The cultural aspect serves both
to atract audience (very few people would show up if all they advertised was
debate) but also as a platform to discuss the issues. By renting a space such
as a local bar or discothèque, they will atract a young crowd who will come for
the music and the entertainment. However, the entertainment will not be empty
entertainment and the evening will be punctuated by other discussion events. In
fact, there is no strict distinction between what Kufunana does to atract a
crowd and what is part of their message, but rather a combination of both
elements in different fractions.

In our interviews, the members of Kufunana often blamed ``ignorance as well as
lack of knowledge.'' We realized that they would employ the word ``ignorance''
(actually, the Portuguese equivalent, \textit{ignor\^{a}ncia}, which, in
European Portuguese has the same meaning as the English word) to mean
``choosing to ignore the facts even if one knows them.'' This is one
justification for using theatre for social goals. It is not only that a
performance can attract more audience than a lecture, it can lead to reflection
on behavior and behavioral change while a lecture can only impart knowledge,
which is not always the limiting factor in affecting change.

\chapter{A Rede (The Network)}
La Ronde by Arthur Schnitzler is a play from 1900, set in Vienna. It is
structured as a series of encounters between a man and a woman. Character~1
meets (and has sex) with character~2, which then seen with character~3, which
is seen with character~4. This continues until character~10 has a sexual
meeting with the first character, closing the loop (hence the name of the
play). In each scene, the actual sexual act is represented by a short blackout.
The characters are extracted from all of \textit{fin de siecle} Austrian
society, prostitutes, artists, noblemen,\ldots

One of the readings of the play is that the subtext is the transmission of
Syphilis, the venereal disease of the time, through society. After having
learned about Mozambican society through discussions with the members of
Kufunana and others, we were reminded of Schnitzler's play. We imagined we
could adapt it to address the spread of HIV, the relevant sexually transmitted
disease today.

We described the basic outline of Schnitzler's La Ronde to the elements of
Kufunana. We made it clear that we did not intend to perform this European play
(even if it is a marvelous play), but to use it as set of initial concepts for
discussion and asked for feedback. Some of the elements of Kufunana related it
to a recent concerted effort by HIV-related NGOs to make people aware of the
sexual networks of which they are a part of. Thus the idea developed from the
circle to the network.

We started work to develop these ideas into a show. All of our scenes emerged
organically through improvisation with the actors. Actors were not assigned
roles at this stage and some themes (for example, teacher and female student
exchanging sex for grades) were played by several actors. The actors had a
mandate to explore Mozambican themes and characters, but, beyond that, were
free to improvise. We noted that there was little sense of artistic property
and actors were happy to let others take over the scene they had created.

In addition to on-stage work, we had discussions with the actors on whether the
piece represented Mozambican society and what was missing. For example, at one
point, we realized that we had a rich man, but no politically powerful figures
(particularly important in a society so marred by corruption). This discussion
led us to add a member of parliament.

We had an imbalance in the distribution, with more male actors than female
actors. We had not chosen the actors to work with, but worked with those that
were interested. Partially for this reason, but also to introduce outside
observers that could comment on the actions of the other characters, we
introduced what we called a ``Greek chorus of drunkards.'' Their goals and
actions were envisioned to be similar to the Greek chorus of classical
tragedies: to foreshadow events, to tell the back story, and to provide the
\textit{vox popolis}. However, to introduce a Greek chorus would be a
completely foreign element to Mozambican culture and would not resonate with
the audience. Therefore, we had the men of the chorus sit at a bar, on one end
of the stage, drinking and talking about the main actions between the scenes.
Their inebriation served as the naturalistic justification for their frankness
about everyone in the community.

For representing two of the sexual scenes, we developed a dance representation
of the sexual transmission of HIV. In one of these scenes, there is viral
transmission. In the other, one of the characters, between the prostitute and a
truck-driver, the prostitute, who, as a professional, always protects herself,
wears a condom. Thus, she is able to fight off the infection. Again, even if
the scene is meaningful artistically and socially, it emerged from discussions
where we learned that, today, prostitutes do tend to protect themselves.  The
scene was developed with the actors and set to traditional African drumming.

\bigskip
\bigskip
\bigskip

In summary, we used a collaborative effort to adapt a century old play about
\textit{fin de siecle} Vienna to modern day Mozambique. As outsiders, we
brought this idea to the group, but let their responses organically mold the
show as we would be otherwise unable to create something that spoke to a
community with which we were unfamiliar. Although the content was particular to
the Mozambican society, we think our working approach would be useful in any
community in which one is working with as an outsider with local actors. As a
side effect of creating this show, we learned much about the structure of
Mozambican society, particularly its sexual economy, that the actors would
portrait on stage and which we would then follow up with discussions.

In Mozambique, sex is not often given for free, but in exchange for gifts or
money (even inside of stable relationships). One of the couples in the play, a
young man and woman, were set to represent an exchange-free relationship.
Although we were assured that this does happen in their society, at one point,
one of the actors had to resorted to explaining to another the point of that
scene: ``it's like in the movies, they are together for love, not for anything
else.''

From La Ronde, we kept the structure that almost all of the scenes to be
partner scenes with a sexual encounter towards the end. We had characters from
all strata of society. Most of the characters in Schnitzler's play find a
correspondant in our version: the teacher for the poet, the congressman and his
wife for the Husband and Young Wife, the policeman for the soldier\ldots One
character is present without any cultural adaptation, namely, the prostitute.

Artistically, the result is sometimes a mix of Western and local, for example
the danced scenes would not be out of place in a modern Western show, but they
were set to traditional African drumming. Another scene, set in a discotheque,
featured loud Mozambican pop, local music inspired by Western forms.

\chapter{Conclusions}

We have described our interactions with Kufunana and how they use theatre to
reach out. We have left out some of their other activities as they would be
beyond the scope of this paper, but we will note that they perform many other
services for the community.

We described our work with them, which we believe can serve as a model for
working in any community to which one comes as an outsider. We brought our
inputs, our techniques, but we used them to start a discussion and then used
mainly improvisation on stage and discussion off stage to mold the show into
something that could be used to discuss the issues in that society.

\end{document}
